---
title: "Basic R Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

<style>

.tt {
  font-family:'Lucida Console', monospace; 
  text-align:center
}

h2 {
  color: darkgreen;
  font-weight: bold
}

h3 {
  color: darkolivegreen;
}

.topicsContainer {
  background-color: #f2f7f6;

}

.topic.current {
  background-color: #e6fcf8;
  border-color: #41ab98;
  border: solid 1px;
  font-weight: bold
}

.band {
  padding-left: 0%
}

.topicsFooter .resetButton {
  font-size: 80%;
}

.btn-primary, .btn-success, .btn-info {
    background-color: darkgreen;
    background-image: none;
    border: none;
}
    .btn-primary:hover, .btn-primary:focus, .btn-primary:active, .btn-primary:focus:active, .btn-success:hover, .btn-success:focus, .btn-success:active, .btn-success:focus:active, .btn-info:hover, .btn-info:focus, .btn-info:active, .btn-info:focus:active {
    background-color: darkolivegreen; }
  .btn-primary:focus:hover, .btn-primary:active:hover, .btn-primary:focus:active:hover, .btn-success:focus:hover, .btn-success:active:hover, .btn-success:focus:active:hover, .btn-info:focus:hover, .btn-info:active:hover, .btn-info:focus:active:hover {
    background-color: darkolivegreen; }
  .btn-primary.disabled, .btn-primary:disabled, .btn-primary.disabled:hover, .btn-primary:disabled:hover, .btn-success.disabled, .btn-success:disabled, .btn-success.disabled:hover, .btn-success:disabled:hover, .btn-info.disabled, .btn-info:disabled, .btn-info.disabled:hover, .btn-info:disabled:hover {
    background-color: darkolivegreen; }
    
.boxed{
 border: 3px solid green ;
 padding: 5px;
}


</style>


## Creating Vectors

We can create vectors using an assignment arrow and the <span class="tt">c()</span> function such as


<p class = "tt">
name <- c(element1, element2, ... , elementn)
</p>

Note that names in R :

* Are case sensitive
* May contain numbers, letters, and .'s but must start with a letter
* Should be descriptive wherever possible
* Are often written in camel case such as getMean or firstName when they need two words

For example, to make a vector with the first five even positive numbers integers, we would write:

```{r, echo = TRUE}
z <- c(2, 4, 6, 8, 10)
z
```

<div class = "boxed">
**NOTE:** We wrote <span class = "tt">z <- c(2, 4, 6, 8, 10)</span> to create the vector but then wrote the name of the vector so that it would **display** the contents of <span class = "tt">z</span>
</div>

### Exercise 

Write the R code required to create a vector called <span class="tt">x</span> that contains the numbers 2, 4, 6, 8, 10 and then display the contents of <span class="tt">x</span>:

```{r make-x, exercise=TRUE}

```

```{r make-x-solution}
x <- c(2, 4, 6, 8, 10)
x
```

To create a vector with a sequence of integers from $i$ to $j$, we can use a :

<p class = "tt"> z <- c(i:j) <p>

would create a vector named <span class ="tt">z</span> that contained the numbers $i$, $i+1$, $i+2$, up to $j$.

For example, to create a vector containing the numbers 1 to 5 we would go:

```{r, echo = TRUE}
z <- c(1:5)
z
```

### Exercise

Write the R code required to create a vector called <span class="tt"> y </span> that contains the integers from 7 to 11 and the display the contents of <span class="tt">y</span>.

```{r make-y, exercise=TRUE}

```

```{r make-y-solution}
y <- c(7:11)
y
```

## Math with Vectors

### Arithmetic on a Single Vector

If we would like to perform an arithmetic operation to **every** element of a vector, we can use basic math functions such +, -, *, and / for addition, subtraction, multiplication, and division. For e.g.

```{r, echo = TRUE}
z <- c(1, 4, 7, 9)
z*2
```

Notice this took each element of <span class ="tt"> z </span> and multiplied every one by 2. You will also see that R only outputted the result of <span class="tt">z * 2</span> and **not** from <span class="tt">z</span>. This is because <span class="tt">z</span> was created as an assignment and <span class="tt">z * 2 </span> was not. We would have to have a line that just said <span class="tt">z</span> or used the <span class="tt">print()</span> command on <span class="tt">z</span> such as

```{r, echo = TRUE}
z <- c(1, 4, 7, 9)
print(z)
```

```{r, echo = TRUE}
z <- c(1, 4, 7, 9)
z
```

Here are additional examples of arithmetic on vectors:

```{r, echo = TRUE}
z <- c(1:5)
z + 4
```

```{r, echo = TRUE}
z <- c(1:5)
z - 3
```

```{r, echo = TRUE}
z <- c(3, 6, 9, 12, 15)
z / 3
```

```{r, echo = TRUE}
z <- c(1:5)
2 * (z - 1)
```

### Exercise

Create a vector <span class="tt">x</span> that contains the integers from 1 to 6. Use R to calculate the values that are two less than three times the values in <span class="tt">x</span>. Note you will need to combine two arithmetic operations in the correct order.

```{r arithmetic, exercise = TRUE, exercise.lines = 3}

```

```{r arithmetic-solution}
x <- c(1:6)
3 * x - 1
```

### Arithmetic on Multiple Vectors

If we have two (or more) vectors of equal length, we can use arithmetic operations on them to perform math **component-wise**. That means the operations are applied to the first element of each vector, then separetly to the second, and again separetly to the third, etc...

Suppose I had information on how many hours each of my 5 employees work at regular pay and at over time pay (paid at a rate of 1.5 times their normal wage). If they are all paid \$10 an hour, I might calculate their pay as follows:

```{r, echo = TRUE}
regular <- c(6, 9, 5, 4, 3)
overtime <- c(1, 2, 0, 0, 3)
pay <- 10 * regular + 1.5 * 10 * overtime
print(pay)
```

We can see the first employee worked 6 regular hours and 1 overtime hour. They were paid \$10 for 6 hours and \$15 for 1 hour for a total of 10\*6+15\*1=\$75. That is the first element of the computed pay vector. The second employee worked 9 regular hours and 2 overtime hours. They were paid \$10 for 9 hours and \$15 for 2 hours for a total of 10\*9+15\*2=\$120. That is the second element of the computed pay vector.

### Exercise

In the R chunk below I have created three vectors with information on the number of hot dogs, fries, and drinks five customers bought at a hotdog stand. Hotdogs cost \$4, fries cost \$3.50, and drinks cost \$2. Below the provided code, create a vector called <span class="tt">total</span> that contains the total cost of all the orders and <span class="tt">print()</span> the contents of the vector.

```{r hotdog, exercise = TRUE, exercise.lines = 5}
hotdogs <- c(3, 2, 1, 1, 4)
fries <- c(3, 3, 2, 1, 6)
drinks <- c(3, 0, 2, 1, 4)
```

```{r hotdog-solution}
hotdogs <- c(3, 2, 1, 1, 4)
fries <- c(3, 3, 2, 1, 6)
drinks <- c(3, 0, 2, 1, 4)
total <- c(4 * hotdogs + 3.50 * fries + 2 * drinks)
print(total)
```

## Dataframes

### Creating Dataframes

We saw in the last section that we often have multiple pieces of information for each observation in our dataset. For the hot dog example, for each order we tracked the number of hotdogs, the number of fries, and the number of drinks for each individual order. In this case, we may want to store all the information into one object instead of three individual vectors. In R this is called a **dataframe**.

In a dataframe, each column represents a variable and each row represents an observation in our dataset. The columns have names associated with them so that we can call upon a specific column by name later. We can make a dataframe from a series of vectors by going:

<p class = "tt", style = "text-align: center"> name <- dataframe(vector1, vector2, ... vectorn)</p>

Here is an example of a dataframe containing information on 6 students in a small seminar class. It has information on their name, their major, the number of credit hours they've completed, and their grade on their first essay:

```{r, echo = TRUE}
name <- c("Johnson, Amy", "Zhang, Xiao", "Brar, Jasdeep", "Dubois, Genevieve", "Germaine, Sophie", "Bector, Bhupinder")
major <- c("Philosophy", "English", "Political Science", "English", "Philosophy", "Gender Studies")
credits <- c(39, 63, 93, 45, 66, 54)
essay1 <- c(75, 92, 87, 77, 90, 85)
seminar <- data.frame(name, major, credits, essay1)
print(seminar)
```

* Notice that the dataframe was able to hold a mix of both numeric and character variables. If we want R to read data as character data (qualitative), we must put it in quotation marks. 

If you have a csv file that contains data, you can save the file in your working directory (can be found and set in the *Session* drop down menu), and use the <span class = "tt"> read.csv() </span> function. This will save the information in the csv file as a dataframe.

For e.g., if I had the information on the seminar class saved in a file called seminar.csv, I would save the file to my working directory and use the code

```
seminar <- read.csv("seminar.csv")
```

to create my dataframe. If you are unfamiliar with making csv files, you can also make a spreadsheet in a program such as Excel or Sheets and ask it to save your file as a csv file.

### Exercise

Take the information in the following table and write R code to create a dataframe called <span class = "tt">pets </span> containing the same information with the same variable names. Print the contents of the dataframe to make sure it matches.

```{r}
name <- c("Fluffy", "Mittens", "George", "Zoinks", "Nermal")
animal <- c("rabbit", "cat", "dog", "mouse", "cat")
age <- c(3, 5, 2, 1, 7)
pets <- data.frame(name, animal, age)
knitr::kable(pets)
```

```{r pets, exercise = TRUE, exercise.lines = 5}

```

```{r pets-solution}
name <- c("Fluffy", "Mittens", "George", "Zoinks", "Nermal")
animal <- c("rabbit", "cat", "dog", "mouse", "cat")
age <- c(3, 5, 2, 1, 7)
pets <- data.frame(name, animal, age)
print(pets)
```

### Accessing Information in a Dataframe

When we are working with a dataframe, it is possible that we want to look/compute values on a particular row, a particular column, or a particular entry. When I know the column that I want to use, I can for it using the format:

<p span = "tt"> dataframeName$columnName </p>

```{r columns-setup}
name <- c("Johnson, Amy", "Zhang, Xiao", "Brar, Jasdeep", "Dubois, Genevieve", "Germaine, Sophie", "Bector, Bhupinder")
major <- c("Philosophy", "English", "Political Science", "English", "Philosophy", "Gender Studies")
credits <- c(39, 63, 93, 45, 66, 54)
essay1 <- c(75, 92, 87, 77, 90, 85)
seminar <- data.frame(name, major, credits, essay1)
```

Recall our seminar class information dataframe called seminar.

```{r, exercise.setup = "columns-setup"}
print(seminar)
```

Suppose I wanted a list of all the grades on the first essay, I could call for that information in the dataframe by going:

```{r, exercise.setup = "columns-setup", echo = TRUE}
seminar$essay1
```

I can also apply a function directly to that column. For e.g., in R, the function <span class ="tt">mean()</span> finds the average of a vector of numbers. I could find the average score on the first essay by going:

```{r, exercise.setup = "columns-setup", echo = TRUE}
mean(seminar$essay1)
```

We can also access particular rows, columns, or cells by attaching two square brackets next to our dataframe's name and giving a locator in the form:

<p class = "tt">name[row, column]<p>

In our <span class = "tt">seminar</span> data frame, the credits are listed in the third column. To get a list of the credit hours we can go

```{r, exercise.setup = "columns-setup", echo = TRUE}
seminar[,3]
```

Notice we only gave a number for the column but still included the comma to make sure R knew we were asking for column 3 instead of row 3.

If we wanted information on Genevieve who appears in the fourth row (we exclude the column names from the rows), we can go

```{r, exercise.setup = "columns-setup", echo = TRUE}
seminar[4,]
```

If we specifically want to know Jasdeep's major, we note that Jasdeep is in row **3** and the major is listed in column **2** so we can go

```{r, exercise.setup = "columns-setup", echo = TRUE}
seminar[3,2]
```

### Quiz

```{r quiz-setup}
Character <- c("Ziggy", "Roberto", "Jasmine", "Jericho", "Herbert")
Class <- c("Rogue", "Priest", "Mage", "Barbarian", "Bard")
Gold <- c(90, 170, 220, 110, 200)
Level <- c(6, 8, 8, 7, 5)
rpg <- data.frame(Character, Class, Gold, Level)
```

Andrea is dungeon mastering a role playing campaign with her friends. She has tracked information on all of the players characters: the **Character** name, their **Class**, the amount of **Gold** they have, and their player **Level**. All of this data has been entered into a dataframe in R. Here is a printout of her dataframe named **rpg**:

```{r, echo = TRUE, exercise.setup = "quiz-setup"}
print(rpg)
```

```{r quiz}
quiz(
  question("Which of these lines of code would tell us the class of the character Roberto?",
           answer("rpg[3,2]"),
           answer("rpg[2,3]"),
           answer("rpg[2,2]", correct = TRUE),
           answer("rpg[3,3]")
           ),
    question("Which of these lines of code would give us information on the gold for all of the characters?",
           answer("rpg[,3]", correct = TRUE),
           answer("rpg[3,]"),
           answer("rpg[,4]", correct = TRUE),
           answer("rpg[4,]")
           )
)
```
<!-- ### Quiz -->

<!-- Some questions using basic arithmetic functions in R: -->

<!-- ```{r quiz} -->
<!-- quiz( -->
<!--   question("Which package contains functions for installing other R packages?", -->
<!--     answer("base"), -->
<!--     answer("tools"), -->
<!--     answer("utils", correct = TRUE), -->
<!--     answer("codetools") -->
<!--   ), -->
<!--   question("Which of the R packages listed below are used to create plots?", -->
<!--     answer("lattice", correct = TRUE), -->
<!--     answer("tools"), -->
<!--     answer("stats"), -->
<!--     answer("grid", correct = TRUE) -->
<!--   ) -->
<!-- ) -->
<!-- ``` -->

